---
title: "Named experimental designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Named experimental designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

While named experimental designs can be muddling to understanding the experimental structure, it is still convenient to be able to concisely describe common structures. 

```{r}
library(edibble)
```

In edibble, some named experimental designs can be constructed using the `make_classical` function.  The classical named designs that are supported now is:

```{r}
find_classical_designs()
```

The output contains information about the design, the code to generate the design using edibble that can be easily copy-and-pasted and the output data frame. If you want the edibble code alone then you can just use `code_classical`.

```{r}

make_classical("crd", n = 30, t = 5)

code_classical("rcbd", t = 4, n = 40)
```

The terminal output has color. You can see the terminal output below for the Split-plot design.

```{r, include = FALSE}
options(cli.num_colors = 256L)
options(asciicast_knitr_svg = TRUE)
asciicast::init_knitr_engine(
  startup = quote({
      library(edibble)
    }),
  echo = TRUE,
  echo_input = FALSE)
```


```{asciicast split-plot-print}
make_classical("split", t1 = 4, t2  = 2, r = 4)
```



